Simulation-guided Data Assimilation for Kinetics Modeling of Potassium Channel Isoforms in Mouse Cardiomyocytes

### 2021-11-11 writing/draft.tex
Writing backup

\begin{abstract}
Electrical signaling plays vital roles in the heart that initiates and controls the synchronous contraction of cardiac muscles. Voltage-gated ion channels (VGICs) are the functional unit of electrical activities in cardiomyocytes. Among major cardiac VGICs, K\textsuperscript{+} channels (K\textsubscript{v}) have a distinctive feature that they have multiple isoforms. Although each K\textsubscript{v} isoform produces a unique but slightly overlapping current, having different functionality, only the sum of these individual K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be recorded through in-vitro experiments. Hence, there is an urgent need to develop a data assimilation method that combines mathematical models with experimental data, I\textsubscript{Ksum} recordings, to estimate individual K\textsuperscript{+} currents. Exponential fitting is a traditional method based on the assumption that prominent K\textsuperscript{+} currents have the shape of an exponential decay function. It has been proven that exponential fitting describes I\textsubscript{Ksum} well with estimated K\textsuperscript{+} currents. However, the mathematical model of this classical method only characterizes the shape of the currents, not the kinetics of corresponding K\textsubscript{v} isoforms. In addition, exponential fitting handles I\textsubscript{Ksum} recordings from the same cell with different conditions independently, so it lacks the cellular-level analysis. This paper presents a data assimilation method based on the cellular-level simulation of the underlying gating kinetics of K\textsubscript{v} isoforms. Computer models are designed with parameters that allow us to control the behavior of the ion channel models. A fractional factorial design is utilized to identify the parameters having significant effects on the current shape. Then, various non-linear optimization methods were applied and compared to calibrate the selected parameters to minimize the discrepancies between the simulated currents and I\textsubscript{Ksum} recordings from the same cell. Further, we develop a graphical user interface (GUI) to provide a user-friendly environment to test the proposed method for practitioners unfamiliar with coding. We conduct a case study with our in-vitro experimental data investigating the effect of reduced glycosylation on K\textsubscript{v}. Experimental results show that the proposed method has strong potential to be a novel analysis tool to decompose I\textsubscript{Ksum} recordings to estimate not only the current characteristics but also kinetics of K\textsubscript{v} isoforms that gives a better understanding of disease-related perturbations.

@@

\section{Introduction}
Electrical signaling plays a vital role in the heart conduction system that controls electrical excitation propagation through the heart, which triggers and synchronizes the contraction of a myriad of cardiac muscle cells \cite{veeraraghavan2014mechanisms}. In turn, this electrical stimulation is responsible for the heart to pump blood throughout the body properly. Cardiac voltage-gated ion channels (VGICs) are membrane proteins in a cardiomyocyte that control specific ions to flow through the membrane \cite{bezanilla2005vgic}. They are fundamental functional units of electrical activities in cardiomyocytes as the electrical excitation results from the orchestrated dynamics of different types of ionic currents controlled by the gating kinetics of VGICs as shown in Figure~\ref{fig:variety_ionc} \cite{grant2009cardiac}. This electrical simulations is called the action potentials (AP), which is the net change of membrane potential, and there are three major cardiac VIGCs contributing to the shape and length of the AP: Na\textsuperscript{+}  (Na\textsubscript{v}), Ca\textsuperscript{2+}  (Ca\textsubscript{v}), and K\textsuperscript{+}  (K\textsubscript{v}) channels. Na\textsubscript{v} are responsible for AP initiation (depolarization phase), Ca\textsubscript{v} are responsible for the prolonged depolarization phase (``plateau'') particularly in larger species like human, while various isoforms of K\textsubscript{v}s are collectively responsible for AP deactivation (repolarization phase). 
\begin{figure}[!ht]
    \centering
    \includegraphics[scale=1]{figures/variety_currents.pdf}
    \caption{A schematic illustration of action potentials in mouse ventricular cardiomyocytes and underlying ionic currents of Na\textsubscript{v}, Ca\textsubscript{v}, and K\textsubscript{v} isoforms \cite{nerbonne2003molecular}.}
    \label{fig:variety_ionc}
\end{figure}

In contrast to Na\textsubscript{v} and Ca\textsubscript{v}, K\textsubscript{v} have a distinctive feature that they have multiple isoforms. Each isoform produces different but slightly overlapping component of the total K\textsuperscript{+} current (I\textsubscript{Ksum}) and contributes to different portions of the AP repolarization. To be specific, as shown in Figure~\ref{fig:variety_ionc}, there are various K\textsubscript{v} isoforms and currents they produce: the transient outward K\textsuperscript{+} current I\textsubscript{Kto} (K\textsubscript{v}4.2/K\textsubscript{v}4.3) several components of delayed rectifier K\textsuperscript{+} currents including I\textsubscript{Ks} (K\textsubscript{v}7.1 and minK), I\textsubscript{Kr} (mERG), I\textsubscript{Kslow1} (K\textsubscript{v}1.5), and I\textsubscript{Kslow2} (K\textsubscript{v}2.1), non-inactivating steady-state K\textsuperscript{+} current I\textsubscript{Kss} (K\textsubscript{2P}), and time independent K\textsuperscript{+} current I\textsubscript{K1} (Kir2 family) \cite{brouillette2004functional, liu2011dissection}. Note that although there are two types of transient outward currents, I\textsubscript{Ktof} and I\textsubscript{Ktos}, only one of them is presented and referred to as I\textsubscript{Kto} because I\textsubscript{Ktof} is only found in ventricular apex myocytes, while I\textsubscript{Ktos} is reported to appear only in the septum \cite{xu1999four}. There are several K\textsubscript{v} isoforms whose exact roles in shaping the cardiac AP are still subject of intensive research. Although it is highly desirable to study each unique K\textsuperscript{+} current rigorously, only I\textsubscript{Ksum} can be recorded during in-vitro experiments.

Hence, we need a data analysis tool to estimate individual K\textsuperscript{+} currents from I\textsubscript{Ksum}. Data assimilation is a scientific method that combines mathematical models and experimental data to calibrate the models; thereby, the in-silico models are compatible with the in-vitro observations. To be specific, for our case, a data assimilation method is required that couples mathematical models of K\textsuperscript{+} models with experimental data of I\textsubscript{Ksum} using whole-cell patch-clamp recordings to estimate various K\textsuperscript{+} currents. Exponential fitting is a traditional method based on the assumption that prominent K\textsuperscript{+} currents from patch-clamp recordings such as I\textsubscript{Kto}, I\textsubscript{Kslow1}, and I\textsubscript{Kslow2} have the shape of an exponential decay function $A_{i}e^{-t/\tau_{i}}$ as in Equation~\ref{eq:sumexp_model}, where $t$ is time variable, and $A_{i}$ and $\tau_{i}$ are the shape parameters. I\textsubscript{kss} is simplified as constant $A\textsubscript{Kss}$. It has been proven that exponential fitting accurately describes I\textsubscript{Ksum} with estimated K\textsuperscript{+} currents and has served as a standard method to analyze the sum of K\textsuperscript{+} currents \cite{brunet2004heterogeneous, liu2011dissection}. However, the mathematical model of this classical method only characterizes the shape of the currents, not the kinetics of corresponding K\textsubscript{v} isoforms. In addition, exponential fitting handles I\textsubscript{Ksum} recordings from the same cell with different conditions independently, so it lacks systematic analyses at the cellular level. 
\begin{equation}
    \text{I}\textsubscript{Ksum} = \sum_{i} A_{i} e^{-t/\tau_{i}} + A\textsubscript{Kss}, \; i \in I \subseteq \{\text{Kto, } \text{Kslow1, } \text{Kslow2}\} \label{eq:sumexp_model}
\end{equation}

On the other hand, computer models of cardiac electrophysiology can simulate complex dynamics of electrical activities at the cellular level that enable the quantitative understanding of biophysical functions such as gating kinetics of VGICs in the behavior of cardiomyocytes \cite{winslow2011integrative}. Therefore, this paper proposes a data assimilation method based on the computer models of K\textsubscript{v} isoforms that facilitates kinetics modeling at the cellular level. K\textsubscript{v} models are developed on biophysical principles of VGICs that are encoded in terms of mathematical equations. These equations have parameters that act like ``knobs'', allowing to control the behavior of the models. Tunning computer models to couple with experimental data requires trial-and-error investigations and domain knowledge in both the kinetics of cardiomyocytes and mathematics of computer models. Since modeling of cardiac electrophysiology is an interdisciplinary field that incorporates biology and engineering, calibration of computer models and interpretation of modeling results can be challenging for researchers in biology or medicine communities without engineering training. This hurdle hinders researchers in physiology from utilizing computer models to analyze their experimental data. A graphical user interface (GUI) is provided in which researchers can easily calibrate K\textsubscript{v} models to their data and visualize the modeling results. A new model calibration method will allow researchers to utilize computer models to analyze complex potassium dynamics in cardiomyocytes without engineering background. A GUI will provide an interactive environment to use the calibration method and analyze the results easily with visualizations.

\textbf{add a paragraph for case study with glycosylation and heart failure}

Cardiac electrophysiology has provided insights into the relationships between the complex dynamics of electrical signaling in the heart at the molecular level and the behavior of cardiomyocytes. Voltage-gated ion channels (VGICs) are the functional unit of electrical activities in cardiomyocytes. Among major cardiac VGICs, potassium channels have a distinctive feature in that they have multiple isoforms. Although each potassium isoform produces a unique current having different functionality, only the sum of these individual potassium currents (IKsum) can be recorded through in-vitro experiments. Hence there is an urgent need to develop a data analysis tool to estimate individual potassium currents from IKsum. Data assimilation is a scientific method that combines mathematical models and experimental data to calibrate the models; thereby, the in-silico models are compatible with the in-vitro observations. Exponential fitting is a traditional method based on the assumption that prominent potassium currents have the shape of an exponential decay function. However, the mathematical model of this classical method only characterizes the shape of the currents, not the kinetics of corresponding potassium channel isoforms. In addition, exponential fitting handles IKsum recordings from the same cell with different conditions independently, so it lacks systematic analyses at the cellular level. Therefore, this paper proposes a data assimilation method based on the computer models of potassium channel isoforms that facilitate kinetics modeling at the cellular level. A new data assimilation method is expected to allow researchers to estimate the kinetics of potassium isoforms from experimental IKsum data.

@@

\section{Material and Methods}
\subsection{Procedure}
Least squares method \cite{reali2017optimization}.

\subsection{Data}

@@

\section{Formatting}

\subsection{Tables}
Tables should be inserted at the end of the manuscript. Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted. Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space. These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance. The author will be notified during the typesetting of the final article if this is the case.

>>>>

\begin{abstract}
Electrical signaling plays vital roles in the heart that initiates and controls the synchronous contraction of cardiac muscles. Voltage-gated ion channels (VGICs) are the functional unit of electrical activities in cardiomyocytes. Among major cardiac VGICs, K\textsuperscript{+} channels (K\textsubscript{v}) have a distinctive feature that they have multiple isoforms. Although each K\textsubscript{v} isoform produces a unique but slightly overlapping current, having different functionality, only the sum of these individual K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be recorded through in-vitro experiments. Hence, there is an urgent need to develop a data assimilation method that combines mathematical models with experimental data, I\textsubscript{Ksum} recordings, to estimate individual K\textsuperscript{+} currents. Exponential fitting is a traditional method based on the assumption that prominent K\textsuperscript{+} currents have the shape of an exponential decay function. It has been proven that exponential fitting describes I\textsubscript{Ksum} well with estimated K\textsuperscript{+} currents. However, the mathematical model of this classical method only characterizes the shape of the currents, not the kinetics of corresponding K\textsubscript{v} isoforms. In addition, exponential fitting handles I\textsubscript{Ksum} recordings from the same cell with different conditions independently, so it lacks the cellular-level analysis. This paper presents a data assimilation method based on the cellular-level simulation of the underlying gating kinetics of K\textsubscript{v} isoforms. Computer models are designed with parameters that allow us to control the behavior of the ion channel models. A fractional factorial design is utilized to identify the parameters having significant effects on the current shape. Then, various nonlinear optimization methods were applied and compared to calibrate the selected parameters to minimize the discrepancies between the simulated currents and I\textsubscript{Ksum} recordings from the same cell. Further, we develop a graphical user interface (GUI) to provide a user-friendly environment to test the proposed method for practitioners unfamiliar with coding. We conduct a case study with our in-vitro experimental data investigating the effect of reduced glycosylation on K\textsubscript{v}. Experimental results show that the proposed method has strong potential to be a novel analysis tool to decompose I\textsubscript{Ksum} recordings to estimate not only the current characteristics but also kinetics of K\textsubscript{v} isoforms that gives a better understanding of disease-related perturbations.

@@

\section{Introduction}
Cardiac muscle cells have a conduction system that controls the propagation of electrical excitation, which triggers and synchronizes the mechanical contraction of the network of a large number of cardiomyocytes \cite{veeraraghavan2014mechanisms}. In turn, this electrical signaling system is responsible for the heart to pump blood throughout the body properly. Cardiac voltage-gated ion channels (VGICs) are membrane proteins in a cardiomyocyte that control specific ions to flow through the membrane \cite{bezanilla2005vgic}. They are fundamental functional units of electrical activities in cardiomyocytes as the electrical excitation results from the orchestrated dynamics of different types of ionic currents controlled by the gating kinetics of VGICs as shown in Figure~\ref{fig:variety_ionc} \cite{grant2009cardiac}. This electrical simulation is called the action potentials (AP), which is the net change of membrane potential, and there are three major cardiac VIGCs contributing to the shape and length of the AP: Na\textsuperscript{+}  (Na\textsubscript{v}), Ca\textsuperscript{2+}  (Ca\textsubscript{v}), and K\textsuperscript{+}  (K\textsubscript{v}) channels. Na\textsubscript{v} are responsible for AP initiation (depolarization phase), Ca\textsubscript{v} are responsible for the prolonged depolarization phase (``plateau'') particularly in larger species like human, while various isoforms of K\textsubscript{v}s are collectively responsible for AP deactivation (repolarization phase). 
\begin{figure}[!ht]
    \centering
    \includegraphics[scale=1]{figures/variety_currents.pdf}
    \caption{A schematic illustration of action potentials in mouse ventricular cardiomyocytes and underlying ionic currents of Na\textsubscript{v}, Ca\textsubscript{v}, and K\textsubscript{v} isoforms \cite{nerbonne2003molecular}.}
    \label{fig:variety_ionc}
\end{figure}

In contrast to Na\textsubscript{v} and Ca\textsubscript{v}, K\textsubscript{v} have a distinctive feature that they have multiple isoforms. Each isoform produces different but slightly overlapping component of the total K\textsuperscript{+} current (I\textsubscript{Ksum}) and contributes to different portions of the AP repolarization. To be specific, as shown in Figure~\ref{fig:variety_ionc}, there are various K\textsubscript{v} isoforms and currents conducted through them: the transient outward K\textsuperscript{+} current I\textsubscript{Kto} (K\textsubscript{v}4.2/K\textsubscript{v}4.3) several components of delayed rectifier K\textsuperscript{+} currents including I\textsubscript{Ks} (K\textsubscript{v}7.1 and minK), I\textsubscript{Kr} (mERG), I\textsubscript{Kslow1} (K\textsubscript{v}1.5), and I\textsubscript{Kslow2} (K\textsubscript{v}2.1), non-inactivating steady-state K\textsuperscript{+} current I\textsubscript{Kss} (K\textsubscript{2P}), and time independent K\textsuperscript{+} current I\textsubscript{K1} (Kir2 family) \cite{brouillette2004functional, liu2011dissection}. Note that although there are two types of transient outward currents, I\textsubscript{Ktof} and I\textsubscript{Ktos}, 
they are referred to as one, I\textsubscript{Kto}, because I\textsubscript{Ktos} is reported to appear only in the ventricular septum myocytes \cite{xu1999four}. There are several K\textsubscript{v} isoforms whose exact roles in shaping the cardiac AP are still subject of intensive research. Although it is highly desirable to study each unique K\textsuperscript{+} current rigorously, only I\textsubscript{Ksum} can be recorded through in-vitro experiments.

Hence, there is an urgent need to develop a data analysis tool to estimate individual K\textsuperscript{+} currents from I\textsubscript{Ksum}. Data assimilation is a scientific method that combines mathematical models and experimental data to calibrate the models; thereby, the in-silico models are compatible with the in-vitro observations. To be specific, for our case, a data assimilation method is required that couples mathematical models of K\textsuperscript{+} models with experimental data of I\textsubscript{Ksum} using whole-cell patch-clamp recordings to estimate various K\textsuperscript{+} currents. Exponential fitting is a traditional method based on the assumption that prominent K\textsuperscript{+} currents from patch-clamp recordings such as I\textsubscript{Kto}, I\textsubscript{Kslow1}, and I\textsubscript{Kslow2} have the shape of an exponential decay function $A_{i}e^{-t/\tau_{i}}$ as in Equation~\ref{eq:sumexp_model}, where $t$ is time variable, and $A_{i}$ and $\tau_{i}$ are the shape parameters. I\textsubscript{kss} is simplified as constant $A\textsubscript{Kss}$. It has been proven that exponential fitting accurately describes I\textsubscript{Ksum} with estimated K\textsuperscript{+} currents and has served as a standard method to analyze the sum of K\textsuperscript{+} currents \cite{brunet2004heterogeneous, liu2011dissection}. However, the mathematical model of this classical method only characterizes the shape of the currents, not the kinetics of corresponding K\textsubscript{v} isoforms. In addition, exponential fitting handles I\textsubscript{Ksum} recordings from the same cell with different conditions independently, so it lacks systematic analyses at the cellular level. 
\begin{equation}
    \text{I}\textsubscript{Ksum} = \sum_{i} A_{i} e^{-t/\tau_{i}} + A\textsubscript{Kss}, \; i \in I \subseteq \{\text{Kto, } \text{Kslow1, } \text{Kslow2}\} \label{eq:sumexp_model}
\end{equation}

On the other hand, computer models of cardiac electrophysiology can simulate complex dynamics of electrical activities at the cellular level that enable the quantitative understanding of biophysical functions such as gating kinetics of VGICs in the behavior of cardiomyocytes \cite{winslow2011integrative}. This paper proposes a data assimilation method based on the computer models of K\textsubscript{v} isoforms that facilitates kinetics modeling at the cellular level. Hodgkin-Huxley-type computer models of potassium isoforms are developed with parameters that act like ``knobs'', allowing to control the behavior of the models. The more parameters the models have, the more flexible they are, but surplus parameters increase the risk of overfitting, make interpretation of calibration results harder, and have a chance to acerbate calibration accuracy. We utilize a fractional factorial design to identify parameters that significantly affect the current shape, which is a standard sensitivity analysis tool \cite{du2015statistical, du2017silico}. A calibration process consists of an objective function and an optimization method to minimize that objective. We derive the objective function measuring discrepancies between experimental I\textsubscript{Ksum} recordings and simulated ones for each cell. VGIC kinetics is a nonlinear process \cite{rudy2006computational}, so nonlinear optimization methods are utilized. A graphical user interface (GUI) is provided in which researchers can easily apply the proposed method to their data and visualize the modeling results.

We conduct a case study with

\textbf{impacts}
\begin{enumerate}
    \item contribution 1
    \item contribution 2
\end{enumerate}

@@

\section{Material and Methods}
\subsection{Computer Models}
Hodgkin-Huxley type modeling; current equations; full model equations in appendix

\subsection{Sensitivity Analysis and Calibration Parameters}
Half-normal plots; calibration parameters; selected calibration parameters

\subsection{Model Calibration}
Least squares method with multiple starting points \cite{reali2017optimization}.
\begin{figure}[!ht]
    \centering
    \includegraphics{figures/overall_diagram.pdf}
    \caption{Caption}
    \label{fig:overall}
\end{figure}

\subsection{GUI}
screenshots of GUI

@@
